---
title: "Coming from biomod2"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4.5,
                      dev.args = list(pointsize = 14))
library(timesift)
```

biomod2 fits a set of species distribution algorithms on one table of predictors and combines them.
`timesift` fits the same algorithms, and the representation of the predictors becomes a second axis of
the comparison. When the predictors are a sensor record (hourly soil temperature, daily air
temperature), the table biomod2 starts from is already a reduction of it, to monthly means or
growing-degree-days. Here the algorithms run at every temporal grain the record supports, are scored
on one set of held-out folds, and are combined afterwards.

Each algorithm is one implementation in the compiled core that R and Python both call, so a model
fitted from either language returns the same numbers. The cores reproduce the reference packages
biomod2 itself calls (rpart, randomForest, gbm, xgboost, maxnet, MASS, earth, mda, mgcv), and each is
pinned against the reference's own output in the test fixtures. Nothing is fitted through biomod2 or
through those packages.

## The mapping

Every biomod2 algorithm is a learner constructor. A variant of an algorithm is an argument of its
constructor.

| biomod2 | `timesift` | notes |
|---|---|---|
| `CTA` | `tree()` | rpart's rules; `prune = "se_sum"` is biomod2's pruning |
| `RF` | `forest()` | randomForest's defaults |
| `RFd` | `forest(balance = TRUE)` | each tree draws as many units from each class as the smaller class holds |
| `GBM` | `boosting()` | gbm's trees and Newton step |
| `XGBOOST` | `boosting(newton = TRUE)` | xgboost's exact greedy trees; `lambda` and `gamma` are its penalties |
| `MAXNET`, `MAXENT` | `maxnet()` | `classes`, `regmult` and `knots` are maxnet's; the model is the background formulation by default |
| `GLM` | `stepwise(direction = "both", terms = "power", max_terms = Inf)` | `y ~ x + I(x^2)` over every column, searched by AIC as `MASS::stepAIC()` does |
| `GAM` | `additive()` | mgcv's `gam(method = "GCV.Cp")`, `k = 10` |
| `MARS` | `mars()` | earth's forward and pruning passes, refitted as a logistic model |
| `FDA` | `discriminant()` | mda's `fda(method = mars)` with the probit recalibration biomod2 applies |
| `SRE` | `envelope()` | `quantile = 0.025`, as `bm_SRE()` |
| `ANN`, `DNN` | `mlp()` | a torch encoder with a joint head over all responses |

`ANN` and `DNN` map to a learner by architecture, and not to numbers: `mlp()` is a different
network from the one biomod2 fits, `hidden` sets the layer widths, `activation` the nonlinearity
and `dropout` the regularisation, and the training settings live in `train_control()`. nnet
minimises an L2-penalised loss by BFGS and cito by a stochastic optimiser, where `mlp()` uses
AdamW, whose weight decay acts differently, so a fit under the same architecture is a comparable
network and not the same one. biomod2's tuned `ANN` (`size = 5`, `decay = 0.1`, `maxit = 200`) is
`mlp(hidden = 5, epochs = 200, weight_decay = 0.1)`, and its tuned `DNN` (`hidden = c(100, 100)`,
`activation = "selu"`, 150 epochs, batch size 100) is `mlp(hidden = c(100, 100), activation =
"selu", epochs = 150, batch_size = 100)`.

biomod2 ships two option sets, its defaults and a tuned set called `"bigboss"`. Where the two
differ, the learners carry both as `preset = "package"` (the fitting package's own defaults, which
biomod2's default set uses) and `preset = "bigboss"`, so `tree(preset = "bigboss")` is `min_split =
5`, `min_leaf = 5`, `cp = 0.001`, `max_depth = 10` and five inner folds. A setting given explicitly
overrides either preset. `mars()`, `discriminant()` and `additive()` have one set, which both of
biomod2's share.

## One run

The record is simulated: 200 units, a year of readings every six hours, three responses that read
the record through a lagged, weekly-grain mechanism. The true grain is known to be the week, which
makes the comparison below checkable.

```{r data}
sim <- simulate_records(n = 200L, mechanism = "lag", variables = 3L, days = 365L,
                        step_hours = 6, prevalence = 0.25, auc = 0.85, seed = 4L)
targets <- data.frame(unit = rownames(sim$y), sim$y)
str(sim$readings)
```

`models` is a named list, so each candidate reports under the biomod2 name. A learner left open runs
at every grain of `sift`; one given `data =` runs at that representation alone.

```{r models}
models <- list(
  CTA = tree(),
  RF = forest(),
  RFd = forest(balance = TRUE),
  GBM = boosting(),
  XGBOOST = boosting(newton = TRUE),
  MAXNET = maxnet(),
  GLM = stepwise(direction = "both", terms = "power", max_terms = Inf),
  GAM = additive(data = grain("season"), k = 5L),
  MARS = mars(),
  FDA = discriminant(),
  SRE = envelope(data = grain("season"))
)
```

Two of them are pinned to the seasonal grain. An additive model holds `k - 1` coefficients per
column, and the 53 weekly columns of this record would need 478 coefficients for 160 training
units, which `additive()` refuses; four seasonal columns need 17. The envelope shuts out one
presence in twenty at each end of every column, so the more columns, the fewer units fall inside
every band, and a coarse grain is what it is meant for.

```{r fit}
fit <- timesift(
  targets, sim$readings,
  y = starts_with("v"), id = unit, time = time,
  models = models,
  sift = grains("week", "month", "season"),
  ensemble = ensemble("weighted", decay = 1.6, min_score = 0.5),
  resampling = cv(v = 5),
  verbose = FALSE
)
fit$candidates[c("candidate", "grain", "bins", "status")][1:6, ]
```

29 candidates were fitted: nine learners at three grains and the two pinned ones at one. All of them
were cross-validated on the same five outer folds and scored on the same cells, so their means share
a denominator.

```{r scores}
fit
```

The AUC of each candidate, the number of responses it scored highest on, and the held-out score of
the procedure that chose among them and of the weighted combination are in the report above. The
four highest means belong to weekly candidates, the grain the simulation puts the signal at. Where
a biomod2 run reads off which algorithm wins on one table, this run also reads off the grain at
which each algorithm does best, and the ordering across grains for one algorithm does not have to
follow the ordering across algorithms.

```{r plot-fit, fig.alt = "Mean AUC of each candidate against the representation it read."}
plot(fit)
```

## Side by side

A biomod2 run formats the data, fits the algorithms under a cross-validation strategy and then
builds the ensembles. The calls below are the same run in each package, over one table of
predictors for biomod2 and over a record for `timesift`. The biomod2 chunk is not evaluated.

```{r biomod2-side, eval = FALSE}
fmt <- BIOMOD_FormatingData(resp.var = y, expl.var = predictors,
                            resp.xy = xy, resp.name = "sp1")
mod <- BIOMOD_Modeling(fmt, modeling.id = "all",
                       models = c("CTA", "RF", "GBM", "XGBOOST", "MAXNET", "GLM",
                                  "GAM", "MARS", "FDA", "SRE"),
                       CV.strategy = "kfold", CV.nb.rep = 1, CV.k = 5,
                       metric.eval = c("TSS", "ROC"))
ens <- BIOMOD_EnsembleModeling(mod, models.chosen = "all",
                               em.algo = c("EMmean", "EMwmean", "EMca"),
                               metric.select = "TSS", metric.select.thresh = 0.5,
                               metric.eval = c("TSS", "ROC"))
```

```{r timesift-side, eval = FALSE}
fit <- timesift(targets, series, y = starts_with("sp"), id = plot_id, time = datetime,
                models = list(CTA = tree(), RF = forest(), GBM = boosting(),
                              XGBOOST = boosting(newton = TRUE), MAXNET = maxnet(),
                              GLM = stepwise(direction = "both", terms = "power",
                                             max_terms = Inf),
                              MARS = mars(), FDA = discriminant()),
                sift = grains("week", "month", "season"),
                ensemble = ensemble("weighted", min_score = 0.5),
                resampling = cv(v = 5))
```

The Python call takes the same constructors and the same arguments:

```python
import timesift as ts

fit = ts.timesift(targets, series, y="sp_*", id="plot_id", time="datetime",
                  models=[ts.tree(), ts.forest(), ts.boosting(newton=True), ts.maxnet(),
                          ts.mars(), ts.discriminant()],
                  sift=ts.grains("week", "month", "season"),
                  ensemble=ts.ensemble("weighted", min_score=0.5),
                  resampling=ts.cv(v=5))
```

`y` takes every response column at once. biomod2 fits one species per call; here the responses share
one fold map, one set of scorable cells and one set of candidates, and a response is a column of the
result.

## Ensembling

biomod2's ensemble algorithms are options of `ensemble()`:

| biomod2 | `timesift` |
|---|---|
| `EMmean` | `ensemble("mean")` |
| `EMmedian` | `ensemble("median")` |
| `EMwmean` | `ensemble("weighted")`; `decay =` is `EMwmean.decay` |
| `EMca` | `ensemble("committee")`; `rule =` picks how each member's cut is learned |
| `EMcv`, `EMci` | `predict(type = "spread")` |
| `metric.select`, `metric.select.thresh` | `metric =`, `min_score =` |
| none | `ensemble("stack")` |

The stack, `timesift`'s default, fits non-negative weights summing to one on the out-of-fold
predictions alone, minimising the response head's loss over the scorable cells. The weights are
therefore never fitted on a prediction the member made for a unit it had been trained on.

`min_score = 0.5` made a candidate ineligible when its mean AUC was below 0.5, which is the
role `metric.select.thresh` plays. The weights the run settled on:

```{r weights}
round(sort(fit$weights, decreasing = TRUE)[1:6], 3)
```

Committee averaging binarises each member at its own threshold, learned from that member's
out-of-fold predictions, and averages the votes. A smaller run on one grain shows it:

```{r committee}
fit_ca <- timesift(
  targets, sim$readings,
  y = starts_with("v"), id = unit, time = time,
  models = models[c("RF", "GBM", "MAXNET", "MARS", "FDA")],
  sift = grains("week"),
  ensemble = ensemble("committee", rule = "kappa"),
  resampling = cv(v = 5),
  verbose = FALSE
)
subset(fit_ca$estimate, metric %in% c("roc_auc", "tss"), c(arm, metric, score, se))
```

`EMcv` and `EMci` read the members of the ensemble side by side. `type = "spread"` returns the
weighted mean, the standard deviation, the coefficient of variation and an interval for every target
and response:

```{r spread}
sp <- predict(fit, targets[1:5, ], sim$readings, type = "spread")
dimnames(sp)[[3]]
round(sp[1:3, 1, ], 3)
```

On one target per map cell this is an uncertainty map in the way `EMcv` is.

## Scores and thresholds

biomod2 reports TSS at a threshold chosen on the same predictions it scores, and so does most
species distribution code. Where presences are thin that choice inflates TSS. `timesift` reads the
threshold of a candidate from its out-of-fold predictions and reports the expected inflation for a
user's own presence counts with `tss_inflation()`. For the three responses above and five folds the
inflation at two population skills reads:

```{r inflation}
tss_inflation(sim$y, fold_map(sim$y, v = 5), skill = c(0.6, 0.9), replicates = 40)
```

`decision_threshold()` returns the cuts a binary prediction uses, and `rule` picks the criterion:
`"youden"` for TSS, `"kappa"`, `"prevalence"` or `"mpa"`, the minimum predicted area cut that keeps
`perc` of the presences.

biomod2's other evaluation statistics come from `table_metric()`, which cuts the predictions by a
rule and reads the table of decisions against observations:

| biomod2 | `timesift` |
|---|---|
| `TSS`, `ROC`, `KAPPA` | `tss()`, `roc_auc()`, `kappa_score()` |
| `POD`, `POFD`, `FAR`, `SR`, `ACCURACY`, `BIAS`, `OR`, `ORSS`, `CSI`, `ETS` | `table_metric(y, p, "pod")` and the same lower-case names |
| `BOYCE` | `boyce_index()` |
| `MPA` | `decision_threshold(rule = "mpa")` |

biomod2 reads each statistic at the cut that brings that statistic closest to its own optimum on a
grid of 100 cuts. `table_metric()` reads it at the cut of the `rule` given, `"youden"` by default,
so the statistics are comparable at one operating point. Every name is a registered metric, so
`grain_ladder(metric = "csi")` reads it.

## What differs from a biomod2 run

- **Absences.** The learners read the absences they are given. `timesift` draws no pseudo-absences.
  `maxnet()` offers both formulations: the default treats every unit as background, as biomod2's
  `MAXNET` does, and `formulation = "absence"` reads the absences as absences under the response
  head's case weights.
- **Folds.** One fold map is dealt once and read by every candidate and the ensemble. `cv()` deals
  units balanced on a stratifying value, `grouped_cv()` keeps units that share a group on one
  side of each split, and `block_cv()` and `env_cv()` hold out a block of space or of predictor
  space. biomod2's `kfold` is `cv()`, `strat` is `cv(by = )`, `block` is `block_cv()`, `env` is
  `env_cv()` and `user.defined` is a fold map passed as `resampling`.
- **Scorable cells.** A response is scored only on folds whose held-out units hold both classes, and
  only fitted on training units that hold both. The mask follows from the response and the fold map
  and involves no model, so every paired comparison runs on matched cells.
- **Case weights.** The response head sets them, and `positive_weights()` gives the presence-absence
  head's. A learner that cannot use them (the background `maxnet()`, `envelope()`) says so in its
  documentation.
- **Presence-absence only for four algorithms.** `maxnet()`, `envelope()`, `discriminant()` and the
  logistic refit of `mars()` need a binary response. The other learners follow the head's loss.
- **Predictors.** The columns of a representation are the predictors, one per bin and channel, and
  a column of `targets` such as elevation reaches the model where `static` names it.

## When biomod2 is the better tool

A study whose predictors are a stack of environmental rasters, with no record behind them, has
nothing for the representation axis to vary, and biomod2's projection onto rasters, its pseudo-absence
sampling strategies and its response curves are built for that setting. `timesift` predicts rows of
a target table, so a map is a prediction with one target per cell and each cell carrying its own
series at the grain the fit reads. Which grain a record should be read at is the question the
package answers, and a study that does not ask it gains little from switching.
