---
title: "Finding the grain"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4.5,
                      dev.args = list(pointsize = 14))
set.seed(1)
library(timesift)
```

The question `timesift` is built around is at what temporal grain a record should be read for a
given prediction. On field data the answer is unknown, so a method that claims to find it is hard
to check there. `simulate_records()` draws a record whose response depends on it at one named grain,
which turns the comparison into a test: the answer is fixed before anything is fitted, and each tool
of the package can be read against it.

This article plants a response at the weekly grain, runs the comparison over four grains, and reads
the result with the ladder, the paired contrast, the mixed-model contrast, the nested selection and
the occlusion profile. Two further records, planted at the season and at the day, show the ladder
moving with the truth, and a record with no signal at all shows what the same tools report when
there is nothing to find.

## Installation

```{r, eval = FALSE}
# Install from CRAN
install.packages("timesift")

# Or install the development version from GitHub
# install.packages("pak")
pak::pak("gcol33/timesift")
```

```bash
# Install from PyPI
pip install timesift

# with the torch encoders, the contrasts and the plots
pip install "timesift[torch,contrasts,plot]"

# Or install the development version from GitHub
pip install git+https://github.com/gcol33/timesift
```

## A record with a known answer

`simulate_records()` draws, for each unit, a year of readings made of three parts: a seasonal cycle
every unit shares, a constant offset of the unit's own, and an autocorrelated anomaly. Measurement
noise is added on top. Each response variable is driven by a weighted mean of the unit's anomaly
over a short stretch of the year, and the shape of the weights sets the grain. Under
`mechanism = "lag"` the weights decay geometrically over four consecutive weeks from an anchor week,
one anchor per variable, so the driver is an exact linear combination of weekly means.

```{r simulate}
sim <- simulate_records(n = 200L, mechanism = "lag", variables = 8L, days = 365L,
                        step_hours = 6, prevalence = 0.3, auc = 0.9, seed = 2L)
sim
sim$design$anchor
```

The true grain of a mechanism is defined as the coarsest grain at which the driver is still an exact
linear function of the binned record. Day bins nest inside weeks, so a daily representation also
holds everything the driver reads, spread over seven times as many columns. Month bins straddle week
boundaries, and a season bin averages thirteen weeks together, so both lose part of the driver. The
anchors are the positions of each variable's first week among the 53 weekly bins of this record.

`auc = 0.9` is the area under the ROC curve of the driver itself against the response. No fitted
model reaches it: the readings carry noise the driver does not, and the weights have to be
estimated from 200 units. The prevalence is shared by every variable.

The first variable reads four weeks starting at its anchor:

```{r weights}
when <- unique(sim$readings$time)
range(when[sim$weights[, 1] > 0])
```

`targets` carries one row per unit with the eight responses, and `sim$readings` is the long table of
readings that belongs to them.

```{r targets}
targets <- data.frame(unit = rownames(sim$y), sim$y)
str(sim$readings)
```

## One run over four grains

`timesift()` builds each representation in `sift`, fits every learner on each, and scores all of
them on one set of outer folds. The learner here is the elastic net with a shorter penalty path and
a looser stopping tolerance than its defaults, to keep this article quick to build. `year_start`
matches the boundary the simulation located its bins with, which matters for the seasonal grain.
`keep_fits = TRUE` keeps the per-fold models, which the occlusion profile later reads.

```{r fit}
en <- elasticnet(n_inner = 5L, n_lambda = 50L, tol = 1e-6)

fit <- timesift(
  targets, sim$readings,
  y = starts_with("v"), id = unit, time = time,
  learners = en,
  sift = grains("day", "week", "month", "season", year_start = "09-01"),
  resampling = cv(v = 5),
  n_inner = 3L,
  keep_fits = TRUE,
  verbose = FALSE
)
fit
```

The weekly candidate has the highest mean AUC, 0.838, and scored highest on all eight responses. The
daily candidate, which holds the same information over 365 columns, sits at 0.784; the monthly and
seasonal candidates fall further, to 0.704 and 0.550, the last close to what a model with no
information would score. The planted grain is recovered, and the level reached is about 0.06 below
the population ceiling of 0.90.

The procedure rows were computed differently. In each outer training fold the four candidates were
cross-validated again on three inner folds, one was chosen and the stack's weights were fitted, and
only then was the outer test fold predicted. Every one of the five folds chose the weekly candidate,
and the stack put all its weight on it, so the procedure's estimate equals the weekly candidate's
outer mean here. That equality is a property of this run, where the choice never wavered; on a
flatter profile the two numbers separate, as the record with no signal below shows.

```{r plot-fit, fig.alt = "Mean AUC of the elastic net at the day, week, month and season grains, peaking at week, with the ensemble level drawn across."}
plot(fit)
```

The estimate is reported under every registered metric. Four of them are the ones a presence-absence
study usually reads:

```{r estimate}
est <- fit$estimate
est[est$arm == "selected" & est$metric %in% c("roc_auc", "tss", "average_precision", "boyce"),
    c("metric", "score", "lower", "upper", "n_variable")]
```

The interval is across the eight responses, which are fitted and scored on the same units and the
same folds, so it carries the spread between responses and not the error they share.

`resampling = cv(v = 5)` dealt the 200 units into five folds of equal size, balanced on richness,
the number of responses present at a unit, because one fold map serves all eight responses at once.
The mask of scorable cells is computed from the responses and that map alone, before any model is
fitted.

```{r folds}
table(fit$folds)
fit$cells
```

All 40 cells are scorable: at a prevalence near 0.3 every fold holds both presences and absences of
every response. Rarer responses change that, as the last section shows.

## The ladder on prebuilt arrays

Where the arrays already exist, `grain_ladder()` runs the same comparison without the run around
it: every learner at every grain, on one fold map and one mask of scorable cells. `grain_matrix()`
builds the four arrays from the readings, and passing the run's own folds makes the ladder
reproduce the run's candidate scores.

```{r ladder}
x <- grain_matrix(sim$readings, unit, time, reading,
                  grain = c("day", "week", "month", "season"), year_start = "09-01")
vapply(x, function(m) dim(m)[2L], numeric(1L))

lad <- grain_ladder(x, sim$y, en, folds = fit$folds, verbose = FALSE)
lad
```

The 160 scorable cells are eight responses in five folds at four grains. The means match the
candidate table above to every printed digit, because the arrays, the learner, the folds and the
mask are the same.

The plot draws each grain's mean with a 95% interval across responses and circles the best grain.

```{r plot-ladder, fig.alt = "Ladder of mean AUC across day, week, month and season, with intervals across responses and an open circle at week."}
plot(lad)
```

The curve rises from day to week and falls from week on. The rise is the cost of variance: the
daily array holds the driver exactly, but the penalised fit has to find it among 365 columns with
160 training units. The fall is the cost of bias, where the bins average away part of what the
driver reads.

## Two arms on matched cells

A step of the ladder is read with `paired_contrast()`. It takes the difference between two arms
inside each cell both scored, averages it within a response over the folds, and summarises the
per-response means. The responses are the replicates: the interval is a Student's t interval across
them, and `p_value` is a Wilcoxon signed-rank test of the per-response differences against zero.

```{r contrast}
rbind(paired_contrast(lad, "week|elasticnet", "month|elasticnet"),
      paired_contrast(lad, "week|elasticnet", "day|elasticnet"))
```

Week leads month by 0.134 and day by 0.054 in AUC, and both intervals exclude zero. Each contrast
rests on 40 paired cells, and every one of the eight responses favours the weekly arm (`n_favour`).
Both p-values are 0.0078, and they are equal because with eight responses all on one side, the
exact signed-rank test has reached the smallest value it can give. That floor depends only on the
number of responses:

```{r p-floor}
n_responses <- 3:10
data.frame(n_responses, smallest_p = 2 / 2^n_responses)
```

With five responses or fewer, no contrast can reach 0.05 however large the difference. The interval
behaves the same way. The same contrast on three of the eight responses, from a ladder at two
grains:

```{r contrast-three}
lad3 <- grain_ladder(x[c("week", "month")], sim$y[, 1:3], en, folds = fit$folds,
                     verbose = FALSE)
paired_contrast(lad3, "week|elasticnet", "month|elasticnet")
```

The difference is of the same sign, but the interval is several times wider, because a t interval on
two degrees of freedom has a large multiplier, and the p-value cannot fall below 0.25.

An arm is named whole, as `"grain|learner"`. A learner named alone would have to take its best
grain, chosen on the held-out scores the contrast is read off, and `paired_contrast()` refuses it.

## The whole curve at once

`grain_contrasts()` fits a mixed model to the per-cell scores of one learner,
`score ~ grain + (1 | variable) + (1 | fold)`, and compares every grain against the best one by
Dunnett's many-to-one procedure. The random effects absorb the differences in level between
responses and between folds, so each comparison is made within a response and within a fold.

```{r grain-contrasts, eval = all(vapply(c("lme4", "lmerTest", "emmeans"), requireNamespace, logical(1), quietly = TRUE))}
grain_contrasts(lad)
```

The singular-fit message from lme4 means that one of the two variance components was estimated at
zero; the fixed-effect contrasts are still read in the usual way. Every grain falls below week, and
the day contrast, the closest, has an adjusted interval that excludes zero. The three intervals have
the same width, about 0.09, because the model estimates one residual variance from all the cells and
the design is balanced. The paired intervals each carry their own spread across responses instead:
the week-to-day one is narrower than its mixed-model counterpart, and the week-to-month one wider.
The reference was itself chosen as the best observed grain, which favours it, so this table is best
read beside the paired contrasts, which single out no grain.

## Choosing inside the training data

The best grain on the ladder was chosen on the same held-out folds its score is read from.
`select_grain()` moves the choice inside the training data. In each outer fold the training units
are split again, every candidate is fitted and scored on the inner folds, the best one is refitted
on the whole outer training set, and the outer test fold is predicted once. The estimate that comes
back is of the procedure, choice included.

```{r select}
sel <- select_grain(x, sim$y, en, folds = fit$folds, n_inner = 3L, verbose = FALSE)
sel
sel$selected[c("fold", "grain", "inner_score", "inner_se")]
```

All five outer folds chose week, and the procedure scores 0.838, the same as the run's estimate,
which made the same choice on the same folds. The inner scores, between 0.767 and 0.817, are lower
than the outer ones because each inner model was fitted on two thirds of 160 units. They are used
only to rank the candidates within a fold.

The plot draws every candidate's inner score in every outer fold, one line per fold, and circles
the candidate each fold chose.

```{r plot-select, fig.height = 6, fig.alt = "Inner AUC of the four candidates in each of five outer folds, every fold circling the weekly candidate."}
plot(sel, ylab = "inner AUC")
```

The five lines have the same shape and every circle sits on the weekly candidate. A selection whose
circles wander between candidates is one whose grid is flat enough for the choice to be arbitrary.

## A record with nothing to find

Under `mechanism = "none"` the driver is drawn independently of the record, so no grain carries
information and every AUC should sit near 0.5.

```{r null}
null <- simulate_records(n = 200L, mechanism = "none", variables = 8L, days = 365L,
                         step_hours = 6, prevalence = 0.3, auc = 0.9, seed = 2L)
x_null <- grain_matrix(null$readings, unit, time, reading,
                       grain = c("day", "week", "month", "season"), year_start = "09-01")
folds_null <- fold_map(null$y, v = 5)
summary(grain_ladder(x_null, null$y, en, folds = folds_null, verbose = FALSE))
```

The ladder has a best grain all the same, the daily one at 0.524. The highest of four means read on
the folds it was chosen on is the number most open to chance, and a reader who took it as the level
would report skill that is not there.

```{r null-select}
select_grain(x_null, null$y, en, folds = folds_null, n_inner = 3L, verbose = FALSE)
```

The procedure scores 0.520 with an interval from 0.480 to 0.560, which contains 0.5. The folds chose
day four times and month once, and the inner scores of all four candidates lie between 0.498 and
0.510. On this record the gap between the best candidate and the procedure is small, 0.004; it grows
with the number of candidates searched and with how much they differ by chance.

`choose = "coarsest_adequate"` replaces the highest inner score by the coarsest candidate within one
standard error of it. On a flat profile that is the least storage the record can be kept at without
a measured loss inside the training data:

```{r null-coarsest}
coarse <- select_grain(x_null, null$y, en, folds = folds_null, n_inner = 3L,
                       choose = "coarsest_adequate", verbose = FALSE)
coarse$selected[c("fold", "grain", "inner_score", "inner_best", "inner_se")]
```

The five folds chose all four grains between them. In folds 2 and 5 the coarsest adequate candidate
was the best one itself; in the others a coarser grain lay within one standard error of the best.
A choice that changes from fold to fold is what a flat profile looks like, and it is the profile to
expect when the record carries nothing the response reads.

## The ladder moves with the truth

Two more records share the design but plant the driver elsewhere. Under `"season"` the weights are
uniform over one season bin, so the true grain is the season and every finer grain nests inside it.
Under `"event"` they are uniform over three consecutive days, so the true grain is the day, and a
week mixes the three days with four others.

```{r mechanisms}
ladder_for <- function(mechanism) {
  s <- simulate_records(n = 200L, mechanism = mechanism, variables = 8L, days = 365L,
                        step_hours = 6, prevalence = 0.3, auc = 0.9, seed = 2L)
  arrays <- grain_matrix(s$readings, unit, time, reading,
                         grain = c("day", "week", "month", "season"), year_start = "09-01")
  grain_ladder(arrays, s$y, en, folds = fold_map(s$y, v = 5), verbose = FALSE)
}
lad_season <- ladder_for("season")
lad_event <- ladder_for("event")
summary(lad_season)[c("grain", "score", "best")]
summary(lad_event)[c("grain", "score", "best")]
```

Each ladder peaks at its planted grain: the seasonal record at season, 0.835, and the event record
at day, 0.864. The seasonal ladder falls gently from season to month, 0.809, where the month array
holds the driver exactly over three times as many columns, and steeply below that. The event ladder
drops by 0.083 from day to week, the information the weekly mean averages away.

```{r plot-event, fig.alt = "Ladder of mean AUC for the event record, highest at day and falling through week, month and season."}
plot(lad_event)
```

## What the weekly model read

`occlusion()` holds one bin back at a time from the per-fold models the run kept, rescores the
held-out units, and records the fall in score as that bin's weight. Nothing is refitted. Here each
weekly bin is permuted across the held-out units five times.

```{r occlusion}
weight <- occlusion(fit, "elasticnet / week", permutations = 5)
weight
```

The planted weights are known, so the profile can be checked response by response. The first
variable's anchor is the second weekly bin, the week of 6 September 2021; the second variable's is
the ninth, the week of 25 October.

```{r occlusion-check}
anchors <- dimnames(x$week)[[2L]][sim$design$anchor[1:2]]
anchors
heaviest <- function(v) {
  w <- as.data.frame(weight)
  w <- w[w$variable == v, c("part", "weight", "importance")]
  head(w[order(-w$weight), ], 4)
}
heaviest("v01")
heaviest("v02")
```

For both variables the heaviest bin is the anchor week, with a fall in AUC of 0.287 and 0.247. The
planted weights decay by a factor of e from one week to the next, and for the first variable the
following week is also among the heaviest. Bins outside the planted stretch carry smaller weights,
such as 0.112 for a January week in the first variable, which suggests the penalised fit picks up
some structure the mechanism does not contain when it has 160 units to fit 53 columns.

`over = "channel"` holds back a whole statistic instead. The simulated response reads the weekly
mean, so a run that also offers each week's coldest and warmest day should lean on the mean.

```{r occlusion-channel}
fit_ch <- timesift(targets, sim$readings, y = starts_with("v"), id = unit, time = time,
                   learners = en,
                   sift = grain("week", stats = c("cold_day", "mean", "warm_day"),
                                year_start = "09-01"),
                   resampling = cv(v = 5), n_inner = NULL, ensemble = FALSE,
                   keep_fits = TRUE, verbose = FALSE)
by_channel <- occlusion(fit_ch, "elasticnet / week", over = "channel", permutations = 5)
aggregate(weight ~ part, by_channel, mean)
```

The mean carries the largest fall, 0.157, about twice the coldest day's. The extreme days are not
independent of the mean, which is why they carry some weight too.

## How the prediction responds

`response_curve()` moves one predictor across its observed range while every other is held at its
mean, and reads the prediction off the model fitted on all targets. A predictor here is one cell of
the representation: the weekly mean in the first variable's anchor week.

```{r response-curve, fig.alt = "Predicted probability against the mean reading in the week of 6 September 2021, rising steeply for v01 and nearly flat for the other responses."}
rc <- response_curve(fit, "elasticnet / week", list(bin = anchors[1], channel = "mean"), n = 20)
plot(rc, col = c("black", rep("grey70", 7)), legend = "topleft")
tapply(rc$prediction, rc$variable, function(p) round(diff(range(p)), 3))
```

The first variable's prediction rises with that week's mean, as the positive sign of its planted
driver says it should, and spans 0.926 of the probability scale. The other variables read different
weeks. Five of them do not move at the third decimal, a sixth moves by 0.001, and the eighth by
0.097, a coefficient the penalised fit gave that cell without the mechanism putting weight there.

## Scores, thresholds and their inflation

A run scores presence-absence responses by `roc_auc()` unless told otherwise. TSS, the statistic
species distribution studies usually report, is the maximum of sensitivity plus specificity minus
one over every cut, and the cut is chosen on the units the score is then read on. That choice
inflates the level, by more the fewer presences a cell holds. `score_predictions()` scores any matrix
of held-out predictions on the scorable cells, here the weekly candidate's out-of-fold predictions
under four metrics:

```{r score}
p_week <- fit$oof[["elasticnet / week"]]
per_metric <- vapply(c("roc_auc", "tss", "average_precision", "boyce"), function(m) {
  s <- score_predictions(sim$y, p_week, fit$folds, metric = m)
  mean(tapply(s$score, s$variable, mean, na.rm = TRUE))
}, numeric(1L))
round(per_metric, 3)
```

Each metric reads the predictions differently. `average_precision()` has the prevalence as its
floor, here about 0.3, where AUC has 0.5. `boyce_index()` uses no absences and correlates the ratio
of presences to all units with the prediction across its range. `tss()` is the one with a cut to
choose. `tss_inflation()` plants predictions of known skill at this design's cell sizes and reads
them back the way a ladder does:

```{r inflation}
tss_inflation(sim$y, fit$folds, skill = c(0.6, 0.9), replicates = 50)
```

At a true skill of 0.6, the level read back is about 0.1 higher; at 0.9 the inflation is smaller,
0.046, nearer the upper bound of 1. `implied_skill()` inverts the map: given the TSS the weekly
candidate reached, it returns the population skill whose expected reported level that is.

```{r implied}
implied_skill(sim$y, fit$folds, observed = per_metric[["tss"]], replicates = 50)
```

A TSS of 0.625 on this design is consistent with a population skill near 0.51. The inversion holds
under the distribution of predictions `tss_inflation()` plants, and two models of equal skill whose
predictions are distributed differently can be inflated by different amounts. A paired difference in
TSS can therefore favour one arm with no difference in skill behind it, which is why a run is scored
by AUC, which has no cut, and reports TSS beside it.

### Rare responses and the scorable-cell mask

A score needs both classes among the held-out units of a cell. With rare responses some cells hold
no presence, and the mask, computed from the response and the fold map alone, leaves them out for
every arm alike.

```{r rare}
rare <- simulate_records(n = 200L, mechanism = "lag", variables = 8L, days = 365L,
                         step_hours = 6, prevalence = 0.03, auc = 0.9, seed = 2L)
colSums(rare$y)
folds_rare <- fold_map(rare$y, v = 5)
scorable_cells(rare$y, folds_rare)
tss_inflation(rare$y, folds_rare, skill = c(0.6, 0.9), replicates = 50)
```

At a prevalence of 0.03 the eight responses hold between 4 and 11 presences in 200 units, eight of
the forty cells are unscorable, and the inflation of TSS at a true skill of 0.6 rises to 0.235. A
response with four presences spread over five folds is scored on at most four of them, each holding
one presence.

## Practical guidance

These points come from the simulations above. They describe these designs, and a study with other
sizes can run the same checks on its own response and fold map.

- The paired contrast treats responses as replicates. With eight responses all favouring one arm,
  its exact p-value was 0.0078, the smallest it can be; with five or fewer no contrast can reach
  0.05, and with three the interval on the week-to-month step was several times wider than with
  eight. A comparison of grains on a handful of species is best reported as differences with their
  intervals, without a test.
- The mixed-model contrast uses every cell and pools one residual variance across grains, so its
  intervals all had the same width, narrower than the paired interval for one step and wider for
  another. It takes the best observed grain as its reference, so it is read beside the paired
  contrasts.
- With 200 units and five outer folds, 40 units are held out per fold. For the lag record that was
  enough to separate week from day by 0.054 AUC with all eight responses agreeing. The selection's
  inner folds see two thirds of each outer training set, so their scores sit lower than the outer
  ones and serve only to rank.
- Rare responses shrink the scorable cells and widen the TSS inflation. `scorable_cells()` and
  `tss_inflation()` run in seconds on the response and the fold map before any model is fitted, and
  fewer folds leave more presences in each held-out cell.
- The best candidate's outer mean is chosen on the folds it is scored on. On the record with no
  signal it read 0.524 while the procedure's interval contained 0.5. The level to report is the
  procedure's estimate, from `timesift()` with `n_inner` set or from `select_grain()`.
- Every run in this article used a single fold map. `cv(repeats = )` runs the comparison again on
  new maps and reads the estimate off all of them, which shows how much of a difference between
  grains belongs to one particular split.
- A selected grain says which representation predicted best on these units. On the lag record the
  daily array holds the driver exactly and still lost to the weekly one, because its information is
  spread over more columns; the grain a model prefers depends on the sample size as well as on the
  mechanism.
