---
title: "Learners"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4.5,
                      dev.args = list(pointsize = 14))
set.seed(1)
library(timesift)
has_torch <- requireNamespace("torch", quietly = TRUE) && torch::torch_is_installed()
```

A learner is the model half of a candidate: a representation says how the record is reduced, and a
learner says what is fitted to the reduction. `timesift` ships fifteen learners, from a penalised
logistic regression to a residual convolutional network. Every one of them goes through the same
pair of functions, the same folds and the same scoring as a learner written for a single study, so
adding a model never means changing the code that fits or scores.

This article lists what ships, what each learner can be handed, how the main ones are set, how the
neural encoders are trained and tuned, and how a learner, a response type or a metric of one's own
is added. It ends with what each learner costs on a small record and where each is the wrong
choice.

The examples use one simulated record throughout: 150 units read every six hours for a year, and
four presence-absence responses that depend on the record through a lagged mechanism acting at the
weekly grain. In this simulation the response reads each unit's departure from its own mean, so a
learner can find the signal only at a grain fine enough to keep that departure.

## Installation

```{r, eval = FALSE}
# Install from CRAN
install.packages("timesift")

# Or install the development version from GitHub
# install.packages("pak")
pak::pak("gcol33/timesift")
```

```bash
# Install from PyPI
pip install timesift

# with the torch encoders, the contrasts and the plots
pip install "timesift[torch,contrasts,plot]"

# Or install the development version from GitHub
pip install git+https://github.com/gcol33/timesift
```

```{r data}
sim <- simulate_records(n = 150L, mechanism = "lag", variables = 4L, days = 365L,
                        step_hours = 6, prevalence = 0.3, auc = 0.85, seed = 4L)
targets <- data.frame(unit = rownames(sim$y), sim$y)
colSums(sim$y)
```

## One interface

A learner is a list carrying a `fit` function, a `predict` function and three declarations: what it
reads, whether it covers the responses jointly or one at a time, and which packages it needs.
Printing one shows the declarations and the settings it carries, with `NULL` where a setting takes
its default at the fit.

```{r print}
elasticnet()
forest(balance = TRUE)
```

`fit` is handed a `[unit, bin, channel]` array and the response matrix of the same units, and
`predict` returns one column per response. A learner that fits one model per response is still
handed the whole response matrix, so nothing above the learner layer needs to know which kind it
was. `c()` binds learners into a set, which `timesift()` takes as `learners`.

```{r set}
c(elasticnet(), forest(), cnn())
```

## The learners that ship

Every shipped learner is registered under its own name, and its declarations can be read off the
constructor:

```{r catalogue}
shipped <- learners()
built <- lapply(shipped, function(nm) getExportedValue("timesift", nm)())
data.frame(learner = shipped,
           reads = vapply(built, `[[`, "", "reads"),
           multi = vapply(built, `[[`, "", "multi"),
           needs = vapply(built, function(l) paste(l$needs, collapse = ", "), ""))
```

Twelve learners read a tabular block and fit one model per response. `mlp()` also reads a tabular
block, flattening the channels, but fits every response jointly through one shared embedding, as
the two convolutional encoders do. Only `cnn()` and `rescnn()` read the bins as a sequence. The
three encoders need `torch`. Every other learner runs on the compiled core the package ships, which
the Python package calls as well, so the two languages fit the same model.

Each compiled learner is written from the published method and is checked in the test fixtures
against the package whose numbers it reproduces. The [biomod2 article](biomod2.html) maps each of
them to the biomod2 algorithm it corresponds to, with the option sets biomod2 uses.

| learner | model | numbers checked against | responses |
|---|---|---|---|
| `elasticnet()` | elastic net over every column and its square, penalty chosen by inner cross-validation | glmnet | every head |
| `linear()` | generalised linear model, terms searched by AIC | MASS `stepAIC()`, `glm()` | every head |
| `tree()` | classification and regression tree, pruned by inner cross-validation | rpart | every head |
| `forest()` | random forest | randomForest | every head |
| `boosting()` | gradient boosted trees, first or second order | gbm, xgboost | every head |
| `maxent()` | maximum entropy over maxnet's feature classes | maxnet | presence-absence |
| `envelope()` | surface range envelope | biomod2 `bm_SRE()` | presence-absence |
| `mars()` | multivariate adaptive regression splines | earth | every head |
| `discriminant()` | flexible discriminant analysis on a MARS basis | mda `fda(method = mars)` | presence-absence |
| `additive()` | additive model of thin plate regression splines | mgcv `gam(method = "GCV.Cp")` | every head |
| `perceptron()` | network of one hidden layer, fitted by BFGS | nnet | every head |
| `hierarchical()` | Bayesian logistic model, optional unit intercepts and spatial field | none | presence-absence |
| `mlp()` | fully connected encoder, joint head | torch | every head |
| `cnn()` | convolutional encoder, joint head | torch | every head |
| `rescnn()` | dilated residual convolutional encoder, joint head | torch | every head |

"Every head" means the learner fits whatever loss the response head names: a logistic model under
presence-absence, a Gaussian one under the continuous, abundance and ordinal heads, and a Poisson
one under the count head. `hierarchical()` has no reference package. It finds the posterior by
Laplace's method and integrates its hyperparameters over a grid around their mode (Rue, Martino and
Chopin 2009), and with `spatial = "nngp"` or `"hsgp"` it adds a Gaussian-process field over the
coordinates that `timesift(coords = )` names.

## What a learner may be handed

A tabular learner reads the bins as a block of predictors, one column per bin of each channel.
Given `native()`, the record unreduced, it would be handed one column per reading, so the pair is
refused. When the learner is pinned to `native()` through `data =`, the refusal comes before the
record is read:

```{r refuse-native, error = TRUE}
timesift(targets, sim$readings, y = starts_with("v"), id = unit, time = time,
         learners = forest(data = native()), verbose = FALSE)
```

Inside a `sift` the same pair is skipped and the run goes on. The candidate table keeps the skipped
pair, with the reason in `note`.

```{r skip-native}
mixed <- timesift(targets, sim$readings, y = starts_with("v"), id = unit, time = time,
                  learners = c(elasticnet(), forest()),
                  sift = c(native(), grain("month")),
                  resampling = cv(v = 5), n_inner = NULL, ensemble = FALSE, verbose = FALSE)
mixed$candidates[c("candidate", "bins", "status")]
```

The native representation of this record has 1460 bins, one per reading. A sequence learner is
refused the opposite case: a representation that gives one row of features, such as
`multigrain()`, or one whose record turns out to hold a single bin. That refusal is shown with the
encoders below.

## Several learners in one run

`learners` takes a set or a named list. A named list reports each candidate under its own name,
which is how two settings of one learner sit side by side. Every learner without a `data =` runs at
every grain of `sift`; `linear()` is pinned to the monthly grain here, and the elastic net is given
a looser stopping tolerance, both to keep the run short. The section on settings shows what the
tolerance changes.

```{r run}
models <- list(
  elasticnet = elasticnet(tol = 1e-4),
  linear = linear(data = grain("month")),
  tree = tree(),
  forest = forest(),
  forest_balanced = forest(balance = TRUE),
  gbm = boosting(),
  xgboost = boosting(method = "xgboost"),
  maxent = maxent()
)
run_time <- system.time(
  fit <- timesift(targets, sim$readings, y = starts_with("v"), id = unit, time = time,
                  learners = models, sift = grains("week", "month"),
                  resampling = cv(v = 5), n_inner = 3L, verbose = FALSE)
)
run_time[["elapsed"]]
```

The run took `r round(run_time[["elapsed"]])` seconds for 15 candidates. With three inner folds in
each of five outer folds, every candidate is fitted 5 x (3 + 1) = 20 times, three inner fits and
one outer fit in each outer fold, and once more on every target.

`summary()` lists the candidates in order of their mean score on the outer folds, with the number of
responses on which each scored highest, and then the held-out score of the procedure that chooses
and weights them.

```{r summary}
summary(fit)
```

Each learner run at both grains scores higher at the weekly one, which is the grain the simulation
acts at. At the monthly grain the tree, both forests and gbm score below 0.5, so they find nothing
there. `maxent / week` has the highest mean, 0.750, and scored highest on all
four responses. The procedure rows are the scores to report: inside each outer training fold the
inner folds chose a candidate and fitted the stack's weights, and the outer test fold then scored
that choice once, 0.741 for the selected candidate and 0.740 for the stack. The highest candidate
mean was picked out on the same folds it is scored on.

The candidate table records how each candidate was built: its learner, its grain, the number of
bins and channels, whether it was fitted, and, for a tuned learner, the settings it chose.

```{r candidates}
head(fit$candidates[c("candidate", "learner", "grain", "bins", "status")], 4)
```

## One learner on one array

`fit_learner()` fits a learner to an array from `grain_matrix()` without the folds and the scoring
around it, and `predict()` on the result takes an array of the same bins and channels. The units
below are split in half once. The two arrays are built from the two halves of the record, over the
same calendar weeks.

```{r one-array}
train <- rownames(sim$y)[1:75]
test <- rownames(sim$y)[76:150]
in_train <- sim$readings$unit %in% train
week_train <- grain_matrix(sim$readings[in_train, ], unit, time, reading, grain = "week")
week_test <- grain_matrix(sim$readings[!in_train, ], unit, time, reading, grain = "week")
f <- fit_learner(elasticnet(), week_train, sim$y[train, ])
f
f$model[c("unfitted", "stopped")]
```

`unfitted` names any response with too few presences to choose a penalty on, which is then
predicted its share among the fitting units, and `stopped` names any response whose penalty path
ended early. Neither applies here. The fit carries the bins it was made on, and an array with other
bins is refused:

```{r one-array-refuse, error = TRUE}
month_train <- grain_matrix(sim$readings[in_train, ], unit, time, reading, grain = "month")
month_test <- grain_matrix(sim$readings[!in_train, ], unit, time, reading, grain = "month")
predict(f, month_test)
```

## Settings of the main learners

The settings below are the ones most often changed; each constructor's documentation lists the rest.

`elasticnet()` fits one penalised model per response over every column and, with `squares = TRUE`,
its square. `alpha` mixes the two penalties, `1` being the lasso and `0` ridge. `s` is where the
penalty path is read: `"lambda.1se"`, the largest penalty within one standard error of the least
held-out deviance, is glmnet's default, `"lambda.min"` is the penalty of least deviance, and a
number is a penalty given directly. `tol` is where the coordinate descent stops; the default leaves
the fit as close to the optimum as glmnet's own default does, and a looser one is faster.

`forest()` grows 500 trees by default, with `mtry` the square root of the column count under
presence-absence. `balance = TRUE` draws as many units of each class into every tree as the smaller
class holds, so a rare response's presences are half of each tree's draw. `boosting()` fits gbm's
first-order trees by default and XGBoost's second-order trees with `method = "xgboost"`, whose
`lambda` and `gamma` are the penalty on a leaf and the least gain a split must make. `maxent()`
builds maxnet's feature classes, the letters of `classes`: `l` linear, `q` quadratic, `p` pairwise
products, `h` hinges and `t` thresholds. Left `NULL`, the classes follow each response's presence
count, as maxnet chooses them.

The function below fits a learner on the training half and returns its AUC on the other half,
averaged over the four responses.

```{r settings}
held_out_auc <- function(l) {
  f <- fit_learner(l, week_train, sim$y[train, ])
  p <- predict(f, week_test)
  mean(vapply(colnames(p), function(v) roc_auc(sim$y[test, v], p[, v]), numeric(1)))
}
variants <- list(
  `elasticnet()` = elasticnet(),
  `elasticnet(s = "lambda.min")` = elasticnet(s = "lambda.min"),
  `elasticnet(alpha = 1)` = elasticnet(alpha = 1),
  `elasticnet(tol = 1e-4)` = elasticnet(tol = 1e-4),
  `forest()` = forest(),
  `forest(balance = TRUE)` = forest(balance = TRUE),
  `boosting()` = boosting(),
  `boosting(method = "xgboost")` = boosting(method = "xgboost"),
  `maxent()` = maxent(),
  `maxent(classes = "lq")` = maxent(classes = "lq")
)
round(vapply(variants, held_out_auc, numeric(1)), 3)
```

On this split the elastic net read at `"lambda.min"` scores 0.680 against 0.631 at the default
`"lambda.1se"`, and the lasso scores 0.608. The looser tolerance gives the same AUC as the default to
three decimals, and the timings at the end of the article show what it saves.
`maxent(classes = "lq")` scores 0.693 against 0.665 for the classes maxnet chooses, and both forests
score close to 0.5 at this grain on 75 units. A single split of 75 units is a noisy comparison, so a
difference of this size is worth checking under `timesift()`'s folds, with the two settings as two
named candidates, before acting on it.

## Neural encoders

The encoders separate architecture from training. `mlp()`, `cnn()` and `rescnn()` take the layer
widths, the kernel and the dropout, and `train_control()` holds every training setting: epochs,
batch size, learning rate, optimiser, schedule, early stopping and device. `train_control()` is the
one place a training setting is defaulted. A run hands its control to every neural learner, and a
training setting given to one learner overrides the run's control for that learner alone, on the
settings it names.

```{r control, eval = has_torch}
train_control(epochs = 5L, device = "cpu")
cnn(channels = c(8L, 16L), kernel = 3L, epochs = 10L)
```

The control marks the settings that were not named as `(default)`, and the learner lists under
`training` the one setting it overrides. The run below trains both encoders at the weekly grain for
a handful of epochs on the processor. The run's control sets five epochs and the convolutional
encoder overrides it with ten. This keeps the example quick and leaves both networks undertrained.

```{r encoders, eval = has_torch}
nets <- timesift(targets, sim$readings, y = starts_with("v"), id = unit, time = time,
                 learners = c(mlp(hidden = c(32L, 16L)),
                              cnn(channels = c(8L, 16L), kernel = 3L, epochs = 10L)),
                 sift = grains("week"), control = train_control(epochs = 5L, device = "cpu"),
                 resampling = cv(v = 3), n_inner = NULL, ensemble = FALSE, verbose = FALSE)
summary(nets)
```

After so few epochs both encoders score close to 0.5, as an undertrained network does. The
`responses` column reads `joint`: one network covers all four responses, and each candidate is still
reported once.

A sequence learner pinned to a representation that gives one row of features is refused before the
record is read:

```{r refuse-sequence, error = TRUE, eval = has_torch}
timesift(targets, sim$readings, y = starts_with("v"), id = unit, time = time,
         learners = cnn(data = multigrain(c("week", "month"))), verbose = FALSE)
```

A fitted encoder holds its weights as plain arrays and the device as the setting it was given, so a
fit saved with `saveRDS()` on a machine with a graphics card predicts after `readRDS()` on one
without.

## Tuning

`tune()` wraps a learner so that, each time it is fitted, it cross-validates a grid of its own
settings on the units it is handed and fits the setting that scored best. In a run each outer fold
tunes on its own training units, so the setting a fold is scored with was never chosen on that
fold's held-out units. A grid names settings the learner carries, and every combination is tried.

```{r tune}
tuned <- tune(forest(trees = 200L), list(mtry = c(2L, 8L), min_node = c(1L, 5L)), n_inner = 3L)
tf <- fit_learner(tuned, week_train, sim$y[train, ])
tf$model$chosen
tf$model$table
```

The search chose `mtry = 8` and `min_node = 5`, the combination with the highest mean inner AUC.
Under `timesift()`, the settings each tuned candidate chose on all targets are in the `settings`
column of the candidate table.

Given no `grid`, `tune()` searches the grid registered under the learner's name. Eight shipped
learners carry the grids that biomod2's `BIOMOD_Tuning()` searches:

```{r tunings}
tunings()
```

`register_tuning()` adds a grid, as a named list or as a function of the learner and the array for a
grid that depends on the data, such as one bounded by the number of columns. The elastic net has no
grid registered; the one below searches its mixing.

```{r register-tuning}
register_tuning("elasticnet", list(alpha = c(0.25, 1)))
te <- fit_learner(tune(elasticnet(), n_inner = 3L), month_train, sim$y[train, ])
te$model$table
```

`alpha = 1`, the lasso, scored higher on the three inner folds, and the tuned learner was fitted
with it. The registration lasts for the session, and replacing a grid already registered takes
`overwrite = TRUE`.

## A learner of one's own

`learner()` takes a `fit` and a `predict` function, and `fit` is handed what it declares. A `fit`
with an argument called `weights` receives the response head's case weights, one per cell of the
response. One with `head` receives the head itself, whose `loss` says what to fit toward. One with
`control` receives the resolved `train_control()`, and one with `group` receives the grouping the
outer folds keep whole, so that a split drawn inside the fit keeps it too. A `fit` that declares
none of these is fitted unweighted, with nothing but the array and the response.

The learner below reduces every channel to its mean over the bins and fits one generalised linear
model per response, in the family the head's loss names.

```{r own}
channel_means <- function(data = NULL) {
  learner(
    "channel_means",
    fit = function(x, y, head, weights, ...) {
      f <- data.frame(apply(unclass(x), c(1L, 3L), mean))
      family <- switch(head$loss,
                       binary_cross_entropy = stats::quasibinomial(),
                       squared_error = stats::gaussian(),
                       poisson_deviance = stats::quasipoisson())
      lapply(seq_len(ncol(y)), function(j) {
        stats::glm(y[, j] ~ ., data = f, weights = weights[, j], family = family)
      })
    },
    predict = function(model, x) {
      f <- data.frame(apply(unclass(x), c(1L, 3L), mean))
      vapply(model, function(m) stats::predict(m, f, type = "response"), numeric(nrow(f)))
    },
    data = data
  )
}
channel_means()
```

`register_learner()` makes the constructor available by name, beside the shipped learners, and a
registered learner can then be given to `timesift()` or `fit_learner()` as its name alone. A fit made
by a registered learner is saved as the learner's name and settings, so a session that reads the fit
back with `readRDS()` registers the learner again before predicting from it.

```{r register-learner}
register_learner("channel_means", channel_means)
"channel_means" %in% learners()
own <- timesift(targets, sim$readings, y = starts_with("v"), id = unit, time = time,
                learners = c(elasticnet(), channel_means()),
                sift = grains("week", "month"),
                resampling = cv(v = 5), n_inner = NULL, ensemble = FALSE, verbose = FALSE)
summary(own)
```

`channel_means` scores 0.461 at both grains. The mean of a channel over the whole year is nearly
the same number whichever bins it is averaged from, and in this simulation the responses read each
unit's departure from its own level, which a yearly mean averages away. The weekly elastic net,
which reads every bin, scores 0.701.

## Other responses, and other metrics

The response head says what the values are, how each learner fits them and which cells a score is
defined on. Besides `"presence_absence"`, four heads ship:

```{r responses}
responses()
```

A count is fitted under the Poisson deviance. The counts below are drawn from each unit's mean
reading, and the forest fits them with no change to the learner.

```{r counts}
week <- grain_matrix(sim$readings, unit, time, reading, grain = "week")
rate <- exp(0.5 + 0.8 * as.numeric(scale(apply(week, 1L, mean))))
counts <- matrix(rpois(150, rate), ncol = 1L, dimnames = list(rownames(sim$y), "count"))
cf <- fit_learner(forest(), week_train, counts[train, , drop = FALSE], response = "count")
round(head(predict(cf, week_test), 3), 2)
```

`maxent()`, `envelope()`, `discriminant()` and `hierarchical()` need a presence-absence response and
refuse any other head:

```{r counts-refuse, error = TRUE}
fit_learner(maxent(), week_train, counts[train, , drop = FALSE], response = "count")
```

`register_response()` adds a head. Its specification names `prepare`, `activation`, `loss`,
`metric` and `cells`, and optionally `weights`. The activation and the loss are each one of the
three the learners know: `"sigmoid"`, `"identity"` or `"exp"`, and `"binary_cross_entropy"`,
`"squared_error"` or `"poisson_deviance"`. The shipped presence-absence head weights each presence
by the ratio of absences to presences among the fitting units, capped at 50 by default, through
`positive_weights()`. The weight of a presence in each response of this record:

```{r weights}
w <- positive_weights(sim$y)
apply(w, 2L, max)
```

A head with a different cap reaches every learner through one registration. With a cap of 2, the
presences of `v01` and `v03` weigh 2 and those of `v02` and `v04` keep the weights above.

```{r register-response}
register_response("presence_absence_cap2", list(
  prepare = function(y) {
    y <- as.matrix(y)
    storage.mode(y) <- "double"
    if (!all(y %in% c(0, 1))) stop("a presence-absence response is 0/1.", call. = FALSE)
    y
  },
  activation = "sigmoid",
  loss = "binary_cross_entropy",
  metric = "roc_auc",
  cells = function(y, folds) scorable_cells(y, folds),
  weights = function(y, fitting) positive_weights(y, cap = 2, fitting = fitting)
))
```

A metric is a function of the observed values and the predictions of one held-out cell.
`register_metric()` adds one by name, and `timesift(metric = )` then chooses and reports by it. The
Brier score below has its sign reversed, so that a higher score is better, as the choice of a
candidate assumes of every metric.

```{r register-metric}
register_metric("neg_brier", function(y, p) -mean((y - p)^2))
capped <- timesift(targets, sim$readings, y = starts_with("v"), id = unit, time = time,
                   learners = c(elasticnet(), forest()), sift = grains("month"),
                   response = "presence_absence_cap2", metric = "neg_brier",
                   resampling = cv(v = 5), n_inner = NULL, ensemble = FALSE, verbose = FALSE)
summary(capped)
```

The run is chosen and reported by the Brier score under the capped weights. The elastic net, at
-0.239, scored above the forest, at -0.259, and scored highest on all four responses.

## Choosing learners

The cost of a learner depends on how many columns its representation flattens to, and on how much
of its fit is an inner search. The timings below fit each tabular learner once on all 150 units and
four responses at the weekly grain, which flattens to 53 columns.

```{r timings}
timed <- list(elasticnet = elasticnet(), `elasticnet(tol = 1e-4)` = elasticnet(tol = 1e-4),
              linear = linear(), tree = tree(), forest = forest(), boosting = boosting(),
              xgboost = boosting(method = "xgboost"), maxent = maxent(),
              envelope = envelope(), mars = mars(), discriminant = discriminant(),
              perceptron = perceptron(), hierarchical = hierarchical())
seconds <- vapply(timed, function(l) {
  system.time(fit_learner(l, week, sim$y))[["elapsed"]]
}, numeric(1))
sort(round(seconds, 2), decreasing = TRUE)
```

`additive()` is not in that list, because at its default `k = 10` an additive model of 53 columns
needs more coefficients than there are units:

```{r additive-refuse, error = TRUE}
fit_learner(additive(), week, sim$y)
```

At the monthly grain, 12 columns, it fits, and its cost depends on `k`:

```{r additive-month}
month <- grain_matrix(sim$readings, unit, time, reading, grain = "month")
additive_seconds <- c(
  `k = 10` = system.time(fit_learner(additive(), month, sim$y))[["elapsed"]],
  `k = 5` = system.time(fit_learner(additive(k = 5L), month, sim$y))[["elapsed"]])
round(additive_seconds, 2)
```

With `k = 10` the fit took `r round(additive_seconds[["k = 10"]], 1)` seconds, and with `k = 5` it
took `r round(additive_seconds[["k = 5"]], 1)` seconds.

The elastic net is the learner to start with. It reads every column without a selection step,
chooses its penalty inside the units it is fitted on, and is the arm the package measures a network
against. At its default tolerance it took `r round(seconds[["elasticnet"]], 1)` seconds here, and
`tol = 1e-4` took `r round(seconds[["elasticnet(tol = 1e-4)"]], 1)`, at the same held-out AUC on the
split above. `linear()` was the slowest at `r round(seconds[["linear"]], 1)` seconds: its two-way
search refits the model for every term it could add or drop at every step, and the weekly grain
offers it 106 terms, each column and its square. Pinned to a coarse grain, or searched forward under
a bound as `linear(select = "forward", terms = "column", max_terms = 3)` does, the search has fewer
steps to take. Every other tabular learner took at most
`r round(max(seconds[!names(seconds) %in% c("linear", "elasticnet")]), 2)` seconds.

What a learner can be fitted on depends on the number of units against the number of columns its
representation flattens to:

- The additive model holds one coefficient for the intercept and `k - 1` for each column, and
  refuses a fit with more coefficients than units. On a small sample it needs a coarse grain or a
  small `k`, and its cost grows with `k`, as the timings above show.
- `linear(select = "none")` fits every term without selecting, and over more columns than units it
  separates any response it is given. It is meant for a coarse grain such as
  `data = grain("season")`.
- The envelope keeps a unit inside only where every column agrees, so each added column leaves
  fewer presences inside. It too is meant for a coarse grain.
- The elastic net, the forest and boosting keep a wide block fittable on few units through the
  penalty, the column draw and the shrinkage, which makes them the first learners to run at a fine
  grain on a small sample.
- `maxent()` builds a hinge per column per knot, so its design grows with the grain. A fit whose
  design would pass `max_design` gigabytes is refused with the size it would have taken, and a
  coarser grain or fewer classes brings it within the limit.
- `perceptron()` reads the columns in their own units unless `standardise = TRUE`, and a record in
  degrees saturates its hidden units sooner the wider its range.

The encoders fit every response through one network, so a rarer response borrows from the common
ones. That pooling is what they are for, and it needs many responses and enough units to train a
network on. In the Schrankogel study, 894 plots and 101 species read hourly for three years, a fully
connected network on the series was level with a penalised logistic model on 188 hand-built
features, and the gain came from convolution reading the series at a coarse grain. A study with few
units and few responses may gain little from the encoders over the elastic net, and pays for the
training.

A learner is the wrong choice where it cannot read the response or the representation. `maxent()`,
`envelope()`, `discriminant()` and `hierarchical()` fit presence-absence alone. A tabular learner
cannot read `native()`, and a sequence learner cannot read a `multigrain()` block or a
representation of one bin. A learner can also be handed a reduction that has already discarded the
signal: `channel_means` above was fitted correctly and scored below 0.5, because a yearly mean keeps
nothing of a weekly departure. No learner recovers what the representation dropped, which is why
`timesift()` compares learners across grains.
