---
title: "Representing a record"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4.5,
                      dev.args = list(pointsize = 14))
set.seed(1)
library(timesift)
```

A learner never reads a sensor record as it was logged. It reads an array of numbers per unit,
and between the record and that array sit three decisions: which stretch of time one number stands
for, what that number summarises, and where the stretches begin. `timesift` names each of them as
an argument, builds the array from them in one compiled core, and records on the array how it was
built.

This article follows a record from the long table it arrives in to the array a learner is handed.
It covers the columns the package reads, the seven named grains and how they follow the calendar,
the seven statistics and the difference between an extreme reading and an extreme day, calendars
passed as functions, gaps and partial bins, the representation objects a run is configured with,
lookbacks anchored on each target, calendar position channels, the tabular layout, and the digest
that ties a representation to its specification.

## Installation

```{r, eval = FALSE}
# Install from CRAN
install.packages("timesift")

# Or install the development version from GitHub
# install.packages("pak")
pak::pak("gcol33/timesift")
```

```bash
# Install from PyPI
pip install timesift

# with the torch encoders, the contrasts and the plots
pip install "timesift[torch,contrasts,plot]"

# Or install the development version from GitHub
pip install git+https://github.com/gcol33/timesift
```

## Two tables, and the columns read from them

A run takes two tables. `series` holds the record in long form, one row per reading, and `targets`
holds one row per thing to predict. They are joined by a column naming the unit, which both carry.
The record here is thirty plots, each logged hourly for 400 days from 1 September 2021, with a
seasonal cycle, a daily cycle, a level per plot and noise.

```{r data}
hours <- seq(as.POSIXct("2021-09-01", tz = "UTC"), by = "hour", length.out = 24 * 400)
ids <- sprintf("p%02d", 1:30)
warmth <- rnorm(30)
doy <- as.numeric(format(hours, "%j"))
hod <- as.numeric(format(hours, "%H"))

series <- data.frame(
  plot = rep(ids, each = length(hours)),
  t = rep(hours, times = 30),
  temp = as.numeric(vapply(warmth, function(w) {
    w + 8 * sin(2 * pi * (doy - 110) / 365) + 3 * sin(2 * pi * (hod - 9) / 24) +
      rnorm(length(hours), sd = 1.5)
  }, numeric(length(hours))))
)

targets <- data.frame(plot = ids, elevation = round(2400 - 150 * warmth))
targets[paste0("sp", 1:4)] <- lapply(c(1, -1, 1, -1), function(s) {
  rbinom(30, 1, plogis(2 * s * warmth))
})
str(series)
head(targets, 3)
```

`grain_matrix()` reads three columns of `series`: the unit (`id`), the instant of each reading
(`time`) and the reading itself (`value`). Each is given as a bare column name or as a string. The
result is a numeric array of shape `[unit, bin, channel]`, and the attributes carry everything the
binning decided.

```{r first}
x <- grain_matrix(series, plot, t, temp, grain = "month")
x
dim(x)
names(attributes(x))
```

The thirty plots became thirty rows, the 400 days fourteen monthly bins, and the single statistic
one channel. `bin_start` and `bin_end` give each bin's first instant on the calendar and its last
reading, `bin_n` the count of readings every cell was reduced from, and `bin_partial` which bins
the record does not cover in full. Bins are named by their start as an ISO 8601 instant in UTC,
which is how the Python side names them too.

A unit identifier is a name. Text and factor levels are used as they are, and a whole number is
written as its digits, so an identifier read from a file as `100000` stays `100000` and never
becomes `1e+05`. The rows are sorted under C collation, the same order on every machine and in
both languages, which puts `"100000"` before `"7"`.

```{r ids}
numeric_ids <- transform(series[series$plot %in% c("p01", "p02"), ],
                         plot = ifelse(plot == "p01", 100000, 7))
dimnames(grain_matrix(numeric_ids, plot, t, temp, grain = "month"))[[1]]
```

A number that is not whole is refused as an identifier, as is a missing value in the unit or the
time column. A unit holding two readings at the same instant is refused as well, with the first
duplicate named, because a reading counted twice would weigh twice in every mean and every daily
extreme it reaches.

```{r duplicate, error = TRUE}
grain_matrix(rbind(series[1:5, ], series[3, ]), plot, t, temp, grain = "day")
```

`timesift()` reads the same columns through its own `id` and `time` arguments, and `x` selects
which numeric columns of `series` hold readings; by default every numeric column but the
identifier and the time. Where it selects more than one, each reaches the learner as channels of
its own, named by the column and the statistic, such as `temp_mean` beside `snow_mean`.

### Time zones

`time` must be `POSIXct`. Bins follow the calendar the column is carried in, which is its
`tzone` attribute, and a column without one is read as UTC. The zone is resolved once, as the
column enters the core, and everything below works on the local clock, so a day is a wall-clock
day. In a zone that keeps summer time, the day the clock goes forward holds 23 hourly readings.

```{r zones}
spring <- seq(as.POSIXct("2021-03-26", tz = "Europe/Vienna"),
              as.POSIXct("2021-03-30 23:00", tz = "Europe/Vienna"), by = "hour")
vienna <- data.frame(plot = "a", t = spring, temp = seq_along(spring))
attr(grain_matrix(vienna, plot, t, temp, grain = "day"), "bin_n")

fixed <- vienna
attr(fixed$t, "tzone") <- "Etc/GMT-1"
attr(grain_matrix(fixed, plot, t, temp, grain = "day"), "bin_n")

identical(as.numeric(grain_matrix(vienna, plot, t, temp, grain = "native")),
          as.numeric(grain_matrix(fixed, plot, t, temp, grain = "native")))
```

Carried in `"Europe/Vienna"`, 28 March 2021 holds 23 readings, and the bins after it start at
22:00 UTC because local midnight has moved an hour. The same instants carried in `"Etc/GMT-1"`,
the fixed offset of Central European standard time, bin into days of 24 readings; the last day
there holds 23 only because the record stops at 23:00 on the summer clock, which is 22:00 on the
fixed one. Changing the `tzone` attribute changes nothing about the readings, only the calendar
they are binned on. The `native` grain is the one grain not read on a clock: its bin is the
reading itself, so it is the same array in either zone.

A logger that keeps a fixed offset is best carried in that offset. The Schrankogel deposit is
one such record: 26,304 hourly readings that bin into 1,096 days of 24.

## The seven grains

A grain names how much of the record one bin stands for. Naming several at once returns a
`timesift_set`, one array per grain over the same units.

```{r ladder}
ladder <- grain_matrix(series, plot, t, temp,
                       grain = c("native", "halfday", "day", "week", "month", "season", "year"))
ladder
```

`native` is the record unreduced: 9,600 bins of one reading each. `halfday` cuts each calendar day
at noon into 800 bins, and `day` gives 400. The four coarser grains follow the calendar and give
58 weeks, 14 months, 6 seasons and 2 years. None of them is a fixed count of hours.

```{r week-start}
week <- ladder[["week"]]
head(attr(week, "bin_start"), 3)
format(head(attr(week, "bin_start"), 3), "%A")
attr(ladder[["month"]], "bin_n")[1, ]
```

A week is an ISO week, Monday to Sunday, so the first weekly bin begins on Monday 30 August 2021
although the record begins on the Wednesday after it. A month is a calendar month: the counts run
720, 744 and 672 readings for months of 30, 31 and 28 days, and the last month holds 120 readings
because the record stops five days into October 2022.

### Where the year begins

`season` and `year` are counted from `year_start`, a `"MM-DD"` boundary that defaults to the
calendar year. A season is three calendar months counted from that boundary, so the default gives
the quarters January to March, April to June and so on. A hydrological year starting in September
is `"09-01"`, and it moves the seasons with it.

```{r year-start}
hydro <- grain_matrix(series, plot, t, temp, grain = c("season", "year"), year_start = "09-01")
attr(hydro[["season"]], "bin_start")
attr(hydro[["year"]], "bin_n")[1, ]
attr(ladder[["year"]], "bin_n")[1, ]
```

On the September boundary the record is five seasons and two years, of 8,760 and 840 readings. On
the default boundary the same record splits into calendar years of 2,928 and 6,672 readings. Which
boundary fits is a question about the system. The Schrankogel deposit uses a hydrological year from
1 September, and on the calendar year each of its winters falls into two bins.

## Statistics

Each bin is summarised by one or more statistics, one channel each, in the order they are named.
Seven exist, and they fall into two families. `mean`, `min` and `max` act on the readings of the
bin directly. The four day-level statistics first reduce each calendar day of the bin and then
reduce again over the days: `cold_day` and `warm_day` are the coldest and warmest daily mean, and
`mean_daily_min` and `mean_daily_max` are the bin's average daily minimum and maximum.

```{r stats}
stats7 <- c("min", "mean_daily_min", "cold_day", "mean", "warm_day", "mean_daily_max", "max")
m <- grain_matrix(series, plot, t, temp, grain = "month", stats = stats7)
round(m["p01", 1:3, ], 2)
```

Within every bin the two families order themselves around the mean: `min` sits below
`mean_daily_min`, which sits below `mean`, and `min` sits below `cold_day` likewise. In
September 2021 plot `p01` read a coldest hour of -4.61, an average daily minimum of -1.07 and a
coldest day of 1.88, against a mean of 3.72. The figure lays the weekly versions of five of them
over the whole record.

```{r weekly-figure, fig.alt = "Weekly minimum, coldest day, mean, warmest day and maximum of one plot over 400 days, the day-level pair lying inside the reading-level pair."}
wk7 <- grain_matrix(series, plot, t, temp, grain = "week", stats = stats7)
start <- as.Date(attr(wk7, "bin_start"))
shown <- c("min", "cold_day", "mean", "warm_day", "max")
matplot(start, wk7["p01", , shown], type = "l", lty = 1, lwd = c(1, 2, 2, 2, 1),
        col = c("grey60", "steelblue", "black", "firebrick", "grey60"),
        xaxt = "n", xlab = "", ylab = "temperature",
        ylim = range(wk7["p01", , shown]) + c(0, 9))
axis.Date(1, start)
legend("top", legend = shown, lty = 1, lwd = c(1, 2, 2, 2, 1), ncol = 3, bty = "n",
       col = c("grey60", "steelblue", "black", "firebrick", "grey60"))
```

### An extreme hour and an extreme day

`min` is set by a single reading, and a single reading can be a sensor touching a cold stone, a
logger dug up by an animal, or one cold hour before dawn. A day-level statistic reaches that hour
only through its share of that day. The next chunk copies September 2021 of plot `p01` and makes
its coldest hour 20 degrees colder, then reduces both copies.

```{r spike}
one <- series[series$plot == "p01" & series$t < as.POSIXct("2021-10-01", tz = "UTC"), ]
spiked <- one
cold <- which.min(spiked$temp)
spiked$temp[cold] <- spiked$temp[cold] - 20

both <- rbind(transform(one, plot = "as recorded"), transform(spiked, plot = "one hour colder"))
round(grain_matrix(both, plot, t, temp, grain = "month", stats = stats7)[, 1, ], 2)
```

`min` follows the hour all the way, from -4.61 to -24.61. `cold_day` moves from 1.88 to 1.05,
which is the 20 degrees divided over the 24 readings of the day the hour fell on.
`mean_daily_min` moves from -1.07 to -1.74, the 20 degrees divided over the 30 daily minima of the
month, and the mean moves by 0.03. The warm side does not move at all. An extreme day is a state
the unit was in for a whole day; an extreme reading can be one hour.

Where the record is clean, the two still carry different things. In the Schrankogel study a
window's coldest and warmest day carried more than its mean, and more the wider the window: the
gain from adding them grew from 0.006 TSS at a weekly grain to 0.046 at a yearly one.

The day-level statistics need whole days inside a bin, so they are refused at `native` and
`halfday` before any record is read. At `day` they are allowed and coincide with the reading-level
ones: a bin of one day has one daily mean and one daily minimum.

```{r day-only, error = TRUE}
grain_matrix(series, plot, t, temp, grain = "halfday", stats = "cold_day")
```

```{r at-day}
d <- grain_matrix(series, plot, t, temp, grain = "day",
                  stats = c("mean", "cold_day", "min", "mean_daily_min"))
round(d["p01", 1:2, ], 2)
```

## A calendar passed as a function

A calendar the package does not carry is passed as `grain`, as a function of the reading instants
that returns the start of each reading's bin. Seasons cut at the solstices and equinoxes rather than
on the first of a month are the usual case. The function below cuts on 21 March, 21 June,
23 September and 21 December, which are fixed dates close to the astronomical ones.

```{r solstices}
solstice_seasons <- function(when) {
  md <- format(when, "%m-%d", tz = "UTC")
  y <- as.integer(format(when, "%Y", tz = "UTC"))
  k <- (md >= "03-21") + (md >= "06-21") + (md >= "09-23") + (md >= "12-21")
  start <- c("12-21", "03-21", "06-21", "09-23", "12-21")[k + 1L]
  as.POSIXct(paste0(ifelse(k == 0L, y - 1L, y), "-", start), tz = "UTC")
}
s <- grain_matrix(series, plot, t, temp, grain = solstice_seasons,
                  stats = c("cold_day", "mean", "warm_day"))
s
attr(s, "bin_n")[1, ]
attr(s, "bin_partial")
```

The array is the same shape as any other and is reported under the grain `custom`. The first bin
opens on 21 June 2021 and holds the 528 readings from the record's start to 23 September, so it is
marked partial. The last bin holds 312 readings and is not: a supplied calendar declares where its
bins begin but not where the last one was meant to end, so the record's end is taken as its end.
Day-level statistics are allowed on a supplied calendar as long as no bin boundary cuts through a
day, and a calendar that does cut one is refused with the day named.

The Schrankogel study's seasonal rung was of this kind, and its reproduction driver passes the
deposit's astronomical seasons the same way.

## Gaps

Every unit must hold at least one reading in every bin. A logger that lost a week leaves a cell
with nothing to summarise, and `grain_matrix()` refuses the record rather than pad it, because a
padded value would reach the model as if it had been measured. Plot `p07` below
loses the calendar week starting Monday 7 February 2022.

```{r gap, error = TRUE}
lost <- series$plot == "p07" &
  series$t >= as.POSIXct("2022-02-07", tz = "UTC") &
  series$t < as.POSIXct("2022-02-14", tz = "UTC")
gappy <- series[!lost, ]
grain_matrix(gappy, plot, t, temp, grain = "week")
```

`coverage()` lays the same binning out as a count of readings per unit and bin, which is where
the gaps of a refused record are read off. It takes the unit and time columns and a grain, and
no value column, since it counts and does not summarise.

```{r coverage}
cv_week <- coverage(gappy, plot, t, grain = "week")
cv_week
cv_week["p07", 22:26]
coverage(gappy, plot, t, grain = "month")
```

At the weekly grain the table holds one empty cell, in the week of 7 February. At the monthly
grain every unit reaches every bin, because the lost week sits inside February and February still
holds three weeks of `p07`'s readings.

```{r gap-month}
attr(grain_matrix(gappy, plot, t, temp, grain = "month"), "bin_n")[c("p06", "p07"), 5:7]
```

Plot `p07`'s February mean is taken over 504 readings where its neighbours' is over 672. That is
one of the three ways out of a gap: move to a grain the gap does not reach, drop the units that do
not span the record, or cut the record to the span every unit covers. Which one is the analyst's
decision, and `coverage()` is the table it is made on; nothing in the package fills a cell.

A bin that no unit reaches is caught by a second rule: consecutive bins must be one bin apart on
the grain's own calendar. A month missing from every logger would otherwise pass as thirteen
adjacent monthly bins, and a convolution would read January and March as neighbours. `native` and
a supplied calendar are exempt from that rule, since the first has no calendar step and the second
declares its own.

## Partial bins

A bin is partial when the record does not cover its whole calendar span, which happens only at
the two ends of the record. Which bins those are follows from where the record starts and stops
against the calendar.

```{r partial}
attr(ladder[["week"]], "bin_partial")[c(1:2, 57:58)]
which(attr(ladder[["month"]], "bin_partial"))
which(attr(ladder[["season"]], "bin_partial"))
dropped <- grain_matrix(series, plot, t, temp, grain = "month", partial = "drop")
dim(dropped)
```

The record starts on a Wednesday, so the first and last weeks are partial. It starts on the first
of a month, so only the last month, five days of October 2022, is partial. On the default year the
first season opens on 1 July and the last on 1 October, so both are partial too. `bin_partial` is
returned whichever way `partial` is set, so a partial bin that is kept is labelled.

`partial = "keep"`, the default, keeps those bins; `partial = "drop"` removes them, which leaves 13
monthly bins here, and refuses if nothing would be left. Keeping costs a bin whose mean is taken
over fewer readings and whose `cold_day` and `warm_day` are drawn from fewer days, so they sit
closer to the bin's mean than a full bin's would. Dropping costs the ends of the record, up to
three months at each end on a seasonal grain. `timesift()` builds its representations with partial
bins kept.

## Representations as settings

`grain_matrix()` builds an array from a record. A representation describes the array before any
record has been read, so that one object can configure a run, name the candidates it produced, and
rebuild itself for new targets at prediction. There are four constructors.

```{r constructors}
native()
grain("week", stats = c("cold_day", "mean", "warm_day"))
multigrain(c("month", "season"), stats = c("min", "mean", "max"))
lookback("28 days", bins = 4)
lookback("90 days", lag = "30 days")
```

`native()` is `grain("native")`. `grain()` takes one grain name or one calendar function, its
statistics and its `year_start`. `multigrain()` builds several grains and flattens each into one
row per target, side by side, so a column names the grain, the statistic and the bin it came from;
with `grains = NULL` it takes every grain the record supports. `lookback()` reads a fixed span
ending a fixed lag before each target's own instant, cut into `bins` sub-bins.

Each printout says whether the representation is a sequence or a block of features. `native()`,
`grain()` and a lookback of more than one bin are sequences, whose bins are ordered in time and
mean something to a convolution; `multigrain()` and a one-bin lookback are blocks. A learner
declares which it reads, and that decides which pairs a run fits.

The constructors check what they can without a record. A day-level statistic at a sub-day grain is
refused at construction, as are an unknown grain, a duplicated statistic and a malformed
`year_start`.

### Sets of them

`grains()` and `lookbacks()` build sets, which is what a run's `sift` takes. Their arguments are
separate names or one vector, and the settings after them apply to every member.

```{r sets}
grains("day", "week", "month")
grains(c("month", "year"), stats = c("cold_day", "mean", "warm_day"))
grains("auto")
lookbacks("28 days", "84 days", bins = 4)
c(grain("week"), multigrain(c("month", "season")))
```

A set is labelled by each member's own label, and two members that would share a label are
refused unless the list is named. `grains("auto")` stands for every named grain the record gives
at least two bins, in order from `native` to `year`, leaving out any grain the requested statistics
are not defined at. It is resolved when a record is read, which is why its printout names the
statistics and no grains, and why it cannot be combined with named grains through `c()`. There is
no cap on it: on this record it includes `native` at 9,600 bins.

## Lookbacks: several targets on one unit

A calendar grain gives every unit one row. Where a unit carries several targets through time,
such as a plot surveyed in May and again in July, each target needs the record leading up to its
own instant. `lookback_matrix()` reads, for every target, a fixed span ending a fixed lag before
that target's instant. `at` is a data frame with an `id` column naming the unit and an `at` column
of instants, and its row names name the targets.

```{r lookback}
visits <- data.frame(id = c("p01", "p01", "p02"),
                     at = as.POSIXct(c("2022-05-01", "2022-07-01", "2022-07-01"), tz = "UTC"))
rownames(visits) <- c("p01-may", "p01-jul", "p02-jul")
lb <- lookback_matrix(series, plot, t, temp, at = visits, span = "28 days", bins = 4,
                      stats = c("cold_day", "mean", "warm_day"))
lb
dimnames(lb)[[2]]
round(lb[, , "mean"], 2)
```

The rows are the three targets in `at`'s own order, two of them on plot `p01`. The four bins are
named by where each opens relative to the anchor, oldest first, so the columns line up across
targets that share no calendar date. The weekly means before the May visit to `p01` rise from
-2.43 to 0.42; before the July visits they rise from 5.36 to 6.65 on `p01` and from 6.08 to 7.82
on `p02`.

A lookback measures fixed lengths: a month is 30 days and a year 365 there, since every target has
to read the same amount of record for the rows to be comparable. Where the calendar matters,
`grain_matrix()` is the call that follows it. The day-level statistics need each bin to be a whole
number of days and to start on a day boundary, which the anchors at midnight above satisfy.

Every target and bin must hold at least one reading. A bin lying entirely before the start of the
record is refused, with the target and the interval named.

```{r lookback-edge, error = TRUE}
early <- data.frame(id = "p01", at = as.POSIXct("2021-09-15", tz = "UTC"))
lookback_matrix(series, plot, t, temp, at = early, span = "28 days", bins = 4)
```

The target on 15 September asks for four weeks from 18 August, and the record begins on
1 September, so the first two weeks of the lookback hold nothing. `bin_n` on a lookback, as on a
grain, gives the number of readings each cell was reduced from, and is worth reading where an
anchor sits near either end of the record.

## The position of a bin in the year

An encoder that ends in global pooling loses when in the year an event happened.
`calendar_channels()` gives that position back as input: the sine and cosine of each bin's
fractional place in the year, read at the midpoint of the record the bin holds. `bind_channels()`
puts them beside the readings.

```{r channels}
wk <- grain_matrix(series, plot, t, temp, grain = "week", stats = c("cold_day", "mean", "warm_day"))
pos <- calendar_channels(wk)
round(pos["p01", 1:4, ], 3)
both_wk <- bind_channels(wk, pos)
dimnames(both_wk)[[3]]
attr(both_wk, "position")
```

The two channels are identical across units and are a time index, not a summary of the readings,
so whatever a model does with them it could have done with a calendar. The sine and cosine pair is
continuous across the turn of the year, where the fraction itself jumps from one back to zero. The
`position` attribute marks them, and the torch encoders read marked channels at their own
amplitude instead of standardising them.

On a grain finer than a day, `cycles = "day"` adds the place in the daily cycle as a second pair.
At a day or coarser every bin would sit at the same place in the day, and that cycle is refused.

```{r channels-day}
hd <- ladder[["halfday"]]
round(calendar_channels(hd, cycles = c("year", "day"))["p01", 1:4, ], 3)
```

The two halves of each day alternate in sign on the daily pair, since the readings of one are
centred on 05:30 and those of the other on 17:30. A lookback has no place in the calendar, its
bins being placed relative to each target, and is refused.

```{r channels-lookback, error = TRUE}
calendar_channels(lb)
```

`bind_channels()` takes any number of arrays over the same units and bins, keeps the first one's
attributes and joins their channel names. Two channels of the same name are refused. It is also
how a second variable, such as a snow product binned to the same grain, joins the temperature.

## The tabular layout

A sequence learner reads the array as it is. A tabular learner, such as `elasticnet()` or
`forest()`, reads a matrix with one row per unit and one column per predictor. `as.matrix()` on a
representation gives that layout, channel by channel and bin by bin within each.

```{r as-matrix}
tab <- as.matrix(grain_matrix(series, plot, t, temp, grain = "season", stats = c("min", "max")))
dim(tab)
colnames(tab)
```

Two statistics over six seasons give twelve columns, each named `channel@bin`. A channel that
holds the same number in every bin, as a `static` predictor such as elevation does inside a run,
becomes one column named by the channel alone, so a penalised fit sees it once rather than once
per bin. This is the function to call inside a learner of one's own that reads a block.

`feature_matrix()` goes the other way: it turns a table already reduced elsewhere, such as a set
of published bioclimatic summaries, into a `[unit, feature, 1]` array a learner reads like any
other. Units come from the row names or from a leading text column.

```{r feature-matrix}
bio <- data.frame(plot = ids, gdd = round(rnorm(30, 900, 80)), frost_days = rpois(30, 40))
fm <- feature_matrix(bio, label = "bioclim")
fm
head(as.matrix(fm), 3)
```

Each feature is a bin of the one channel, which is why the printout counts two bins. A feature
table carries no time axis, because its reduction has already happened, and that is what makes it
the comparison a grain is measured against.

## Sets of built arrays

`timesift_set()` holds several built arrays over the same targets, keyed by label. It is what
`grain_matrix()` returns for several grains, what `grain_ladder()` fits across, and what a fitted
run keeps as its representations. A grain, a supplied calendar and a feature table can be members
of one set.

```{r set}
set <- timesift_set(list(week = wk, season = s, bioclim = fm))
set
set[c("week", "bioclim")]
```

Every member must cover the same targets in the same order, which is checked on the row names. The
lookback above covers three targets where the weekly array covers thirty plots, so the two cannot
share a set.

```{r set-mismatch, error = TRUE}
timesift_set(list(week = wk, visits = lb))
```

## Digests and the artifacts that cross languages

A representation is a set of numbers fixed by a specification, `inst/spec/representation.md`, and
`digest_array()` is the MD5 of an array written the way that specification defines: every value
to twelve decimal places, unit fastest, then bin, then channel, joined by line feeds. The Python
twin computes the same digest from the same array, and the package's test fixtures pin the digest
of every grain and statistic on a small input, so a change to binning shows up as a changed string.

```{r digest}
digest_array(ladder[["month"]])
digest_array(grain_matrix(series, plot, t, temp, grain = "month"))
```

The month built as part of the seven-grain set and the month built on its own give the same
digest, so they are the same representation to the twelfth decimal place. Comparing two digests
is a quick way to check that a representation built on another machine, in another session or in
Python is the one expected.

The representation is rebuilt in each language from the shared core. The fold map, the response
matrix and the mask of scorable cells are built once and handed over as files, because two random
streams seeded alike in R and Python still draw different folds. `write_folds()`, `read_folds()`
and their siblings for the response and the cells write and read that format: CSV, LF line
endings and rows in C collation, byte for byte the same from either side.

```{r artifacts}
y <- as.matrix(targets[paste0("sp", 1:4)])
rownames(y) <- targets$plot
f <- fold_map(y, v = 3)
path <- tempfile(fileext = ".csv")
write_folds(f, path)
readLines(path, n = 4)
back <- read_folds(path, names(f))
identical(as.integer(back), as.integer(f)) && identical(names(back), names(f))
unlink(path)
```

`read_folds()` aligns by name, never by position: `units` gives the order wanted, a unit the file
lacks is an error, and a unit the file carries beyond them is dropped.

## Into a run

A run reaches the representations through two arguments. `sift` in `timesift()` is the set every
learner without a representation of its own is fitted across, and `data =` on a learner pins that
learner to one representation. The run below uses `grains("auto")` for the default learner and
pins a second elastic net to a `multigrain()` block. It compares the candidates on three folds and
skips the nested estimate and the stack, which keeps it fast.

```{r run}
fit <- timesift(targets, series, y = starts_with("sp"), id = plot, time = t,
                learners = list(elasticnet = elasticnet(),
                                pinned = elasticnet(data = multigrain(c("month", "season")))),
                sift = grains("auto"), resampling = cv(v = 3), n_inner = NULL,
                ensemble = FALSE, verbose = FALSE)
fit$candidates[c("candidate", "representation", "bins", "channels", "status")]
```

`grains("auto")` resolved to all seven named grains, since this record gives each of them at least
two bins. The elastic net reads a block of features, so the pair with `native` was skipped and
listed as not applicable; the other six were fitted. The pinned learner was fitted at its own
representation alone, and its block holds 20 columns, the 14 months and the 6 seasons side by side.
Every candidate was scored on the same folds and cells, so the comparison runs across rows of
different kinds.

With `target_time`, every target is anchored in time and every representation must be a lookback.
`sift` then has to be given as `lookbacks()`, since there is no default set of spans, and a
grouped split keeps all targets of one plot on the same side of each fold.

```{r run-lookback}
anchored <- data.frame(plot = rep(ids, each = 2),
                       when = rep(as.POSIXct(c("2022-05-01", "2022-08-01"), tz = "UTC"), 30))
anchored$sp1 <- rbinom(60, 1, plogis(2 * rep(warmth, each = 2)))
lfit <- timesift(anchored, series, y = sp1, id = plot, time = t, target_time = when,
                 sift = lookbacks("28 days", "84 days", bins = 4),
                 learners = elasticnet(), resampling = grouped_cv("plot", v = 3), n_inner = NULL,
                 ensemble = FALSE, verbose = FALSE)
lfit$candidates[c("candidate", "representation", "bins", "channels", "status")]
```

## Practical guidance

### Which grains a record supports

How many bins a grain gives depends on the length of the record and on where it starts against the
calendar. `coverage()` counts the bins without summarising anything, so it is a cheap way to read
them off for a planned deployment. The function below does that for an hourly record of a given
length from 1 September.

```{r lengths}
span_bins <- function(days, start = "2021-09-01", year_start = "01-01") {
  t <- seq(as.POSIXct(start, tz = "UTC"), by = "hour", length.out = 24 * days)
  one <- data.frame(plot = "a", t = t)
  vapply(c("native", "halfday", "day", "week", "month", "season", "year"), function(g) {
    ncol(coverage(one, plot, t, grain = g, year_start = year_start))
  }, integer(1))
}
rbind(`120 days` = span_bins(120), `1 year` = span_bins(365),
      `3 years` = span_bins(1096, year_start = "09-01"))
```

A record of 120 days gives one year bin, so `grains("auto")` leaves `year` out, and two seasons,
both of them partial. One year from 1 September already gives two calendar years, each partial,
and five seasons. Three years on a September boundary give 36 months, 12 seasons and 3 years with
no partial month, season or year. Every count but the seasons matches the Schrankogel study's
bins; its 13 seasons came from the astronomical calendar passed as a function.

A grain with two bins is in `grains("auto")` but gives a sequence learner little to read, and a
seasonal or yearly grain carried by partial bins at both ends rests on fewer readings than its
name suggests. Reading `bin_n` and `bin_partial` before a long run settles both.

### Memory

The array a grain gives is `units x bins x channels` doubles of eight bytes. For the 894 plots of
the Schrankogel study over three years of hourly readings, the native grain and the weekly
three-channel grain compare as follows, in megabytes.

```{r memory}
plots <- 894
c(native = plots * 26304 * 1, week_three_channels = plots * 157 * 3) * 8 / 2^20
c(native = 26304, week_three_channels = 157 * 3)
```

One channel at `native` is 179 MB for the array alone, before a learner copies it into its own
tensors, and the weekly grain with three channels is under 4 MB: 471 numbers per plot against
26,304. `native()` is meant for sequence learners, and a tabular learner given it is refused
before the array is built. `grains("auto")` builds `native` whenever the record gives two
readings, so on a long record a set named by hand is the cheaper choice.

### Choosing statistics

`mean` is the default and the place to start. Adding `cold_day` and `warm_day` beside it gives a
learner the extremes a unit spent a whole day in, which a single bad reading moves by a
twenty-fourth of its error, and which in the Schrankogel study carried more the coarser the
grain. The study reported its networks on `c("cold_day", "mean", "warm_day")` at the weekly grain. `min` and `max` belong where a single
reading is itself the event, such as a frost hour that kills tissue, and are worth checking against
the record for sensor faults first. `mean_daily_min` and `mean_daily_max` describe a typical day of
the bin and suit questions about daily amplitude. Each statistic is a channel, so three statistics
triple the array and, for a tabular learner, the number of predictors.
